Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-05-03 R/analysis-code/analysis.qmd
Edits to the manuscript and code cleaning

# Load the necessary libraries for the analysis
```{r}
library(here)
library(dplyr)         # for data manipulation tasks like filtering, mutating, summarizing, and arranging
library(lubridate)     # for working with dates and times, offering functions to parse, manipulate, and format date-time objects
library(readr)         # for reading and writing rectangular data in text format, providing faster and more consistent reading of delimited files
library(stargazer)     # for creating LaTeX and ASCII tables for descriptive statistics and regression models
library(mgcv)          # for generalized additive models (GAMs) and other generalized ridge regression models
library(tidyr)         # for tidying and reshaping data for analysis and visualization
library(caret)         # for machine learning tasks like model training, tuning, and evaluation
library(randomForest)  # for building random forest models, a popular machine learning algorithm
library(tidymodels)    # for modeling and machine learning tasks using tidy principles
library(ranger)        # for fast, robust, and scalable random forest algorithms
library(dials)         # for tuning hyperparameters in models
library(vip)           # for variable importance assessment in predictive models
library(ggplot2)       # for creating elegant and customizable visualizations with the grammar of graphics
library(boot)          # for bootstrap resampling and estimating standard errors and confidence intervals
library(yardstick)     # for evaluating predictive models using performance metrics
library(glmnet)        # for fitting generalized linear models with LASSO or ridge regularization
library(mlr3)          # for building, evaluating, and comparing machine learning models
library(data.table)    # for efficient data manipulation and aggregation operations
library(Matrix)        # for sparse and dense matrix operations
library(gbm)           # for gradient boosting machines, a powerful machine learning algorithm
library(knitr)         # for dynamic report generation, integrating R code and output into documents
library(gt)            # for creating beautiful and customizable tables in R Markdown and other formats
library(xgboost)       # for extreme gradient boosting, a scalable and accurate machine learning algorithm)

@@

## 1 Part Random Forest
```{r}
# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

@@

# Define the Random Forest model specification
rf_model_spec_continuous <- rand_forest(trees = 500, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity", seed = 123) %>%
  set_mode("regression")

@@

# Workflow setup
rf_workflow_continuous <- workflow() %>%
  add_recipe(rf_recipe_continuous) %>%
  add_model(rf_model_spec_continuous)

# Tune the model
tuning_results_continuous <- tune_grid(
  rf_workflow_continuous,
  resamples = cv_folds_continuous,
  grid = tuning_grid,
  metrics = metric_set(yardstick::rmse)
)

@@

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  rf_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)

# Extract the fit using the appropriate tidymodels function
fitted_model <- extract_fit_engine(final_fit_continuous)

@@

    # Save the plot
    ggsave("feature_importance_rf.pdf", plot = importance_plot, width = 8, height = 6)
} else {
    cat("The extracted object is not a 'ranger' Random Forest model. Please check the model setup and extraction.\n")
}
print(importance_plot)

### Calculate RMSE for the training and testing sets

# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

@@

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_rf$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_rf$.estimate, "\n")

boot_fit_and_evaluate <- function(data, indices) {
  # Subset the bootstrap sample
  boot_sample <- data[indices, ]

@@

```
### There still seems to be overfitting. The RMSE on the test set is higher than the training set.
### We will try a different model

# Generalized Additive Models (GAMs) with sparsity
### GAMs are flexible models that can capture non-linear relationships between predictors and the outcome.
### Acts similar to LASSO adding a penalty similar to LASSO to enforce sparsity and variable selection
```{r}
library(tidymodels)
library(mgcv)
library(readr) # For read_csv

@@

```
# Gradient Boosting Machine (GBM)
```{r}

library(tidymodels)
library(dplyr)
library(yardstick)
library(ranger) # For Random Forest
library(boot) # For bootstrapping

@@

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  gbm_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)

### Calculate RMSE for the training and testing sets

# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

@@

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_gbm$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_gbm$.estimate, "\n")

# Define a function to fit the model and calculate RMSE
boot_fit_and_evaluate <- function(data, indices) {
  # Extract the bootstrap sample
  bootstrap_data <- data[indices, ]

@@

# SUmmarize the RSME values
```{r}
rmse_table <- data.frame(
  Model = c("Null Model", "Generalized Additive Model", "Random Forest", "LASSO",
            "Generalized Additive Model with Sparsity", "Gradient Boosting Machine"),
  Train_RMSE = c(null_rmse, rmse_train_gam1, train_rmse_result_rf$.estimate, 
                 train_rmse_result_lasso$.estimate, train_rmse_result_GAMS$.estimate, 
                 train_rmse_result_gbm$.estimate),
  Test_RMSE = c(null_rmse, rmse_test_gam1, test_rmse_result_rf$.estimate, 
                test_rmse_result_lasso$.estimate, test_rmse_result_GAMS$.estimate, 
                test_rmse_result_gbm$.estimate)
)
# Sort the table by Test_RMSE column in descending order
rmse_table <- rmse_table[order(-rmse_table$Test_RMSE), ]

>>>>

# Load the necessary libraries for the analysis
```{r}
library(here)
library(dplyr)         # for data manipulation tasks like filtering, mutating, summarizing, and arranging
library(lubridate)     # for working with dates and times, offering functions to parse, manipulate, and format date-time objects
library(readr)         # for reading & writing rectangular data in text format
library(stargazer)     # for creating LaTeX and ASCII tables for descriptive statistics and regression models
library(mgcv)          # for generalized additive models (GAMs) and other generalized ridge regression models
library(tidyr)         # for tidying and reshaping data for analysis and visualization
library(caret)         # for machine learning tasks like model training, tuning, and evaluation
library(randomForest)  # for building random forest models, a popular machine learning algorithm
library(tidymodels)    # for modeling and machine learning tasks using tidy principles
library(ranger)        # for fast, robust, and scalable random forest algorithms
library(dials)         # for tuning hyperparameters in models
library(vip)           # for variable importance assessment in predictive models
library(ggplot2)       # for creating elegant and customizable visualizations with the grammar of graphics
library(boot)          # for bootstrap resampling and estimating standard errors and confidence intervals
library(yardstick)     # for evaluating predictive models using performance metrics
library(glmnet)        # for fitting generalized linear models with LASSO or ridge regularization
library(mlr3)          # for building, evaluating, and comparing machine learning models
library(data.table)    # for efficient data manipulation and aggregation operations
library(Matrix)        # for sparse and dense matrix operations
library(gbm)           # for gradient boosting machines, a powerful machine learning algorithm
library(knitr)         # for dynamic report generation, integrating R code and output into documents
library(gt)            # for creating beautiful and customizable tables in R Markdown and other formats
library(xgboost)       # for extreme gradient boosting, a scalable and accurate machine learning algorithm)

@@

## 1 Part Random Forest
### Load the dataset, define predictors, and prepare the data for modeling.
```{r}
```{r}
# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

@@

# Define cross-validation folds
cv_folds_continuous <- vfold_cv(train_data_continuous, v = 5)
```
### Define the RF model specifications, create recipe, and setup workflow
```{r}
# Define the Random Forest model specification
rf_model_spec_continuous <- rand_forest(trees = 500, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity", seed = 123) %>%
  set_mode("regression")

@@

# Workflow setup
rf_workflow_continuous <- workflow() %>%
  add_recipe(rf_recipe_continuous) %>%
  add_model(rf_model_spec_continuous)
```

## Tune the model, extract the best hyperparameters, and fit the final model
```{r}
```{r}
tuning_results_continuous <- tune_grid(
  rf_workflow_continuous,
  resamples = cv_folds_continuous,
  grid = tuning_grid,
  metrics = metric_set(yardstick::rmse)
)

@@

# View the best hyperparameters
print(best_hps_continuous)

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  rf_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)
```

### Extract the fitted model, calculate feature importance
```{r}
# Extract the fit using the appropriate tidymodels function
fitted_model <- extract_fit_engine(final_fit_continuous)

@@

    # Save the plot
    ggsave("feature_importance_rf.pdf", plot = importance_plot, width = 8, height = 6)
} else {
    cat("The extracted object is not a 'ranger' Random Forest model. Please check the model setup and extraction.\n")
}
print(importance_plot)
```

### Calculate RMSE for the training and testing sets
```{r}
# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

@@

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_rf$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_rf$.estimate, "\n")
```

### Bootstrapping to evaluate model robustness
```{r}
boot_fit_and_evaluate <- function(data, indices) {
  # Subset the bootstrap sample
  boot_sample <- data[indices, ]

@@

```
### There still seems to be overfitting. The RMSE on the test set is higher than the training set.
### Additionally, LASSO models are not ideal for capturing non-linear relationships.
### We will try a different model

# Generalized Additive Models (GAMs) with sparsity
### GAMs are flexible models that can capture non-linear relationships between predictors and the outcome.
### Acts similar to LASSO adding a penalty similar to LASSO to enforce sparsity and variable selection
```{r}
# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

@@

# Gradient Boosting Machine (GBM)
### Gradient Boosting Machine (GBM) is an ensemble learning technique that builds models sequentially, with each new model correcting errors made by the previous ones. It uses decision trees as the base learners and optimizes a loss function, making it effective for both regression and classification tasks.
```{r}
# Load the dataset
mir_1part_all <- read_csv(here("data", "processed-data", "all_mir_clean_joined.csv"))

@@

# Fit the final model with the best hyperparameters on the full training dataset
final_fit_continuous <- fit(
  gbm_workflow_continuous %>% finalize_workflow(best_hps_continuous),
  data = train_data_continuous
)
```

### Calculate RMSE for the training and testing sets
```{r}
# Predict on the training and testing set
train_preds <- predict(final_fit_continuous, new_data = train_data_continuous)$.pred
test_preds <- predict(final_fit_continuous, new_data = test_data_continuous)$.pred

@@

# Display the RMSE results
cat("Training Set RMSE:", train_rmse_result_gbm$.estimate, "\n")
cat("Testing Set RMSE:", test_rmse_result_gbm$.estimate, "\n")
```

### Bootstrap to evaluate GBM robustness
```{r}
# Define a function to fit the model and calculate RMSE
boot_fit_and_evaluate <- function(data, indices) {
  # Extract the bootstrap sample
  bootstrap_data <- data[indices, ]

@@

# Summarize the RSME values
```{r}
rmse_table <- data.frame(
  Model = c("Null Model", "Generalized Additive Model", "Random Forest", "LASSO",
            "Generalized Additive Model with Sparsity", "Gradient Boosting Machine"),
  Train_RMSE = c(null_rmse, rmse_train_gam1, train_rmse_result_rf$.estimate, 
                 train_rmse_result_lasso$.estimate, train_rmse_result_GAMS$.estimate, 
                 train_rmse_result_gbm$.estimate),
  Test_RMSE = c(null_rmse, rmse_test_gam1, test_rmse_result_rf$.estimate, 
                test_rmse_result_lasso$.estimate, test_rmse_result_GAMS$.estimate, 
                test_rmse_result_gbm$.estimate)
)
# Sort the table by Test_RMSE column in descending order
rmse_table <- rmse_table[order(-rmse_table$Test_RMSE), ]

### 2024-05-03 products/manuscript/Manuscript.qmd
Edits to the manuscript and code cleaning

Eastern equine encephalitis (EEE) is a rare but serious disease caused
by an Alphavirus transmitted to humans by infected mosquitoes. With a
fatality rate of 30-50% and potential lifelong disability for survivors,
understanding its transmission cycle is crucial. EEE transmission
involves two cycles: an enzootic cycle between Culiseta melanura
mosquitoes and birds, and an epizootic cycle where bridge vector
mosquitoes transmit the virus to humans and other mammals. While C.
melanura poses minimal direct threat to humans, isolations of EEE virus
in these mosquitoes serve as early indicators of virus circulation.
Conversely, isolations in bridge vector species signal heightened
transmission risk. This research aims to identify significant
relationships between variables to inform dissertation prospectus
development and guide methodological approaches.
:::

# Research question

Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts, and which
models best describe the relationship between these factors?

@@

Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
to humans by the bite of an infected mosquito. Human infections are
rare, but serious [@lindseyEasternEquineEncephalitis2018]. Humans and
other mammals are dead-end hosts for EEE. About 94% of human EEE
infections are asymptomatic. While symptomatic cases are rare, the
consequences can be severe. Mortality rates associated with symptomatic
infections range from 30 to 50% [@morensEasternEquineEncephalitis2019].
Those who survive often experience lifelong disability that includes
debilitating neurological damage
[@massachusettsdepartmentofpublichealthArbovirusSurveillancePlan2024].
On average, there are 11 cases reported in the US annually. Most occur
in Massachusetts and Florida [@vanderkelenRiskIndexModel2014]. Until
recently, nearly all human EEE cases in Massachusetts occurred in two
counties: Bristol and Plymouth.

@@

The Bristol County Mosquito Control Project (BCMCP) coordinates mosquito
surveillance and testing in the county from June to October. BCMCP has
used the same sentinel collection sites and trapping methods for over 40
years. Once trapped, mosquitoes are sorted and counted by species, and
vector species are submitted to the Massachusetts Department of Health’s
(MDPH) State Lab for PCR virus testing. Results are available within 24
hours of submission. When rates of EEE mosquito infections are above a
defined threshold, MDPH notifies the local boards of health and
recommends preventive measures. Recommended measures include outreach
and education to increase personal protective practices . Additionally,
MDPH may recommend that cities and towns in the affected areas cancel
evening outdoor events and discourage outdoor activities when mosquito
vectors are most active (dawn and dusk). When infection rates are high
enough to indicate imminent human transmission, the Commonwealth of
Massachusetts will recommend and fund aerial pesticide applications over
the affected areas.

@@

Bristol County, Massachusetts covers 691 square miles with a population
density of 962 people per square mile. It borders Rhode Island to the
west, Plymouth County, MA to the east, and Norfolk County, MA to the
north. Naushon Island and Martha's Vineyard are across Buzzard's Bay to
the Southeast.

Atlantic White Cedar (AWC) swamps cover more than 4,300 acres across
Massachusetts, with the largest concentrations found in Plymouth and
Bristol counties. These swamps are crucial breeding grounds for Culiseta
melanura (MEL), which is identified as the primary enzootic vector for
Eastern Equine Encephalitis (EEE)
[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007].
Atlantic White Cedar (AWC) trees possess extensive root systems that
create "crypts" capable of retaining water year-round. These crypts are
crucial habitats for the larval growth of Culiseta melanura (MEL) and
also provide shelter during the winter months.

@@

**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts, and which
models best describe the relationship between these factors?** From
prior research we know that the mosquito lifecyle is influenced by a
number of factors. Temperature plays a major role in the mosquito life
cyle virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

While these relationships are well known, they are not well described
for EEE, especially in terms of quantifying the role these two factors
play in EEE transmission. Specifically, there's a gap in the
understanding of how these factors influence the number of EEE vectors
mosquitoes collected in traps and the rate of virus infection among
those mosquitoes.

@@

The mosquito surveillance dataset encompasses counts by species from all
mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The
raw dataset includes over 45,000 records and includes temporal and
spatial variables. The BCMCP, perform mosquito surveillance from June
through October annually.

During this period, CDC light traps, baited with carbon dioxide, are
strategically placed at various sentinel sites weekly. These traps are
designed to attract mosquitoes seeking blood meals and are recognized as
a reliable method for monitoring the actively feeding mosquito
population. Consequently, they are established as the standard approach
for Eastern Equine Encephalitis (EEE) surveillance. Notably, some of
these sentinel sites have been in continuous use for over 50 years.

Additionally, the county deploys gravid traps, which are specifically
aimed at attracting and capturing gravid *Culex* mosquitoes looking for
egg-laying sites. While these traps are primarily used for West Nile
Virus (WNV) surveillance, they hold significant value for EEE monitoring
as well. Given the role of *Culex* species as bridge vectors for EEE,
their capture in gravid traps provides critical insight into the risk of
virus transmission.

The traps are set up in the afternoon and collected the following
morning. Once retrieved, the captured mosquitoes are immediately stored
on dry ice and then in a freezer. This process kills the mosquito and
preserves them. The mosquitoes are sorted by species under a dissecting
scope. Counts for each species are recorded in the database and a
portion of the vector species are driven to the Massachusetts State
Laboratory in Boston. The lab tests each mosquito pool (or batch) for
WNV and EEEv using PCR. Results are shared with BCMCP on the same day of
testing. Virus isolation results are recorded in the collection records.

In addition to mosquito counts by species, the dataset includes the town
or city in which they were collected, the collection date, and whether
the mosquitoes were submitted for virus testing.

### Minimun infection rate (MIR)

In our investigation into the environmental, ecological, and temporal
factors that impact the transmission cycles of EEE in southeastern
Massachusetts, the Mosquito Infection Rate (MIR) emerges as a pivotal
indicator. MIR provides a quantifiable measure of virus prevalence
within mosquito populations, making it an essential metric for modeling
the dynamics of disease spread. By analyzing changes in MIR in response
to varying environmental conditions and mosquito behaviors, we can
enhance our understanding of the key drivers behind EEE transmission.
This understanding, in turn, guides the development of predictive models
that can accurately forecast periods of heightened transmission risk
based on observed or anticipated changes in the identified factors. The
minimum infection rate (MIR) can be calculated at various temporal and
spatial scales from trap site over one day to all sites over an entire
season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ While MIR does not directly measure the risk to humans, a higher MIR
in mosquitoes that commonly bite humans can indicate a higher risk of
transmission. A study pubulihsed in 2007 indicated that in years with
human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by
week 31 (mid-late July) [@hachiyaHumanEasternEquine2007]. MIR allows for
the comparison of infection rates across different geographical areas or
time periods. This is crucial for identifying areas or times of higher
risk and for evaluating the effectiveness of control measures.

@@

The surveillance dataset is structured to conform with state and federal
reporting requirements. As a result, the format of the data must be
transformed to in order to make it ready for analysis. Once the data
structure is corrected, it can be joined to other relevant datasets
using spatial or temporal attributes.

@@

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold. or degree-days. Degree-days is a measure of the amount of
heat accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006]. To simplify this
variable, the number of hours the temperature was ≥50°F and ≤86°F. These
thresholds were chosen based on the known temperature range for mosquito
development and virus transmission.

@@

In response to the observed non-linear relationships in the data, a
two-stage Generalized Additive Model (GAM) analysis was used. GAMs
provide a versatile framework for capturing complex, non-linear
relationships between response variables and multiple predictors, which
are often encountered in ecological data sets.

>>>>

Eastern equine encephalitis (EEE) is a rare but severe disease caused
by an Alphavirus transmitted to humans by infected mosquitoes. With a
fatality rate of 30-50% and potential lifelong disability for survivors,
understanding its transmission cycle is crucial. EEE transmission
involves two cycles: an enzootic cycle between Culiseta melanura
mosquitoes and birds and an epizootic cycle where bridge vector
mosquitoes transmit the virus to humans and other mammals. While *C. melanura* poses minimal direct threat to humans, isolations of EEE virus in these mosquitoes serve as early indicators of virus circulation. In contrast, isolations in bridge vector species signal a heightened transmission risk. This project aims to identify significant relationships between variables to inform the development of a dissertation prospectus and guide methodological approaches..
:::

# Research Question

In southeastern Massachusetts, the transmission dynamics of Eastern Equine Encephalitis (EEE) remain inadequately understood, particularly how environmental, ecological, and temporal factors interplay to affect outbreak patterns. This gap in knowledge hinders effective predictive modeling and response strategies, crucial for preventing and mitigating the impact of EEE.

**Which environmental,ecological, and temporal factors influence the transmission cycles for EEE in southeastern Massachusetts, and which models best describe the relationship between these factors?**

@@

Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
to humans by the bite of an infected mosquito. Human infections are
rare but serious [@lindseyEasternEquineEncephalitis2018]. Humans and
other mammals are dead-end hosts for EEE. About 94% of human EEE
infections are asymptomatic. While symptomatic cases are rare, the
consequences can be severe. Mortality rates associated with symptomatic
cases range from 30 to 50% [@morensEasternEquineEncephalitis2019].
Those who survive often experience lifelong disability that includes
debilitating neurological damage
[@massachusettsdepartmentofpublichealthArbovirusSurveillancePlan2024].
On average, 11 cases reported in the US annually. Most occur
in Massachusetts and Florida [@vanderkelenRiskIndexModel2014]. Until
recently, nearly all human EEE cases in Massachusetts occurred in two
counties: Bristol and Plymouth.

@@

The Bristol County Mosquito Control Project (BCMCP) coordinates mosquito
surveillance and testing in the county from June to October. BCMCP has
used the same sentinel collection sites and trapping methods for over 40
years. Once trapped, mosquitoes are sorted and counted by species, and
vector species are submitted to the Massachusetts Department of Health’s
(MDPH) State Lab for PCR virus testing. Results are available within 24
hours of submission. When rates of EEE mosquito infections are above a
defined threshold, MDPH notifies the local boards of health and
recommends preventive measures. Recommended measures include outreach
and education to increase personal protective practices. Additionally,
MDPH may recommend that cities and towns in the affected areas cancel
evening outdoor events and discourage outdoor activities when mosquito
vectors are most active (dawn and dusk). When infection rates are high
enough to indicate imminent human transmission, the Commonwealth of
Massachusetts will recommend and fund aerial pesticide applications over
the affected areas.

@@

Bristol County, Massachusetts covers 691 square miles with a population
density of 962 people per square mile. It borders Rhode Island to the
west, Plymouth County, MA to the east, and Norfolk County, MA to the
north. Naushon Island and Martha's Vineyard are across Buzzard's Bay.

Atlantic White Cedar (AWC) swamps cover more than 4,300 acres across
Massachusetts, with the largest concentrations found in Plymouth and
Bristol counties. These swamps are crucial breeding grounds for Culiseta
melanura (MEL), which is identified as the primary enzootic vector for
Eastern Equine Encephalitis (EEE)
[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007].
Atlantic White Cedar (AWC) trees possess extensive root systems that
create "crypts" capable of retaining water year-round. These crypts are
crucial habitats for the larval growth of Culiseta melanura (MEL) and
also provide shelter during the winter.

@@

*Which environmental, ecological, and temporal factors influence the transmission cycles for EEE in southeastern Massachusetts, and which models best describe the relationship between these factors?* From prior research, we know that the mosquito lifecycle is influenced by several factors. Temperature plays a significant role in the mosquito life cycle and virus transmission dynamics. As temperature increases, the rate of larval development also increases, and the time between when a mosquito ingests the virus in a blood meal and when it can transmit it decreases. The result is that the mosquito population can grow at a faster rate, and they will have more time to transmit the virus. However, MEL larvae develop in well-insulated crypts that are less sensitive to ambient air temperature changes, so the effect of temperature changes on MEL development speed may be less pronounced than other vector species that develop in open, standing water, like ditches and containers. Precipitation increases the amount of standing water and the chances of larval survival during the mosquito season and winter for those species that overwinter in the larval stage. For EEE, the primary enzootic vector, MEL, overwinters as larvae.
While these relationships are well known, they are not well described for EEE, especially in quantifying the role these two factors play in EEE transmission. Specifically, there’s a gap in understanding how these factors influence the number of EEE vector mosquitoes collected in traps and the rate of virus infection among those mosquitoes

@@

The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The raw dataset includes over 45,000 records and includes temporal and spatial variables. The BCMCP performs mosquito surveillance from June through October annually.
During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites weekly. Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance. Some of the sentinel sites have been used for over 50 years.
Additionally, the county employs gravid traps specifically aimed at attracting and capturing gravid Culex mosquitoes looking for egg-laying sites. While these traps are primarily used for West Nile Virus (WNV) surveillance, they also hold significant value for EEE monitoring. Given the role of Culex species as bridge vectors for EEE, their capture in gravid traps provides critical insight into the risk of virus transmission.
The traps are installed in the afternoon and retrieved the following morning. After the traps are collected, the next are stored in a freezer to preserve and kill the mosquitoes. The mosquitoes are sorted by species using a dissecting scope. Counts for each species are recorded in the database, and a portion of the vector species are driven to the Massachusetts State Laboratory in Boston. The lab tests each mosquito pool (or batch) for WNV and EEEv using PCR. Results are shared with BCMCP on the same day of testing. Virus isolation results are recorded in the collection records.
In addition to mosquito counts by species, the dataset includes the town or city in which they were collected, the collection date, and whether the mosquitoes were submitted for virus testing.

### Minimum infection rate (MIR)

In our investigation into the environmental, ecological, and temporal
factors that impact the transmission cycles of EEE in southeastern
Massachusetts, the Mosquito Infection Rate (MIR) emerges as a pivotal
indicator. MIR provides a quantifiable measure of virus prevalence
within mosquito populations, making it an essential metric for modeling
the dynamics of disease spread. By analyzing changes in MIR in response
to varying environmental conditions and mosquito behaviors, we can
enhance our understanding of the key drivers behind EEE transmission.
This understanding, in turn, guides the development of predictive models
that can accurately forecast periods of heightened transmission risk
based on observed or anticipated changes in the identified factors. The
minimum infection rate (MIR) can be calculated at various temporal and
spatial scales from trap site over one day to all sites over an entire
season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ While MIR does not directly measure the risk to humans, a higher MIR in mosquitoes that commonly bite humans can indicate a higher risk of transmission. A study published in 2007 indicated that in years with human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by week 31 (mid-late July) [@hachiyaHumanEasternEquine2007]. MIR allows for comparing infection rates across different geographical areas or time periods. This is crucial for identifying areas or times of higher risk and evaluating control measures' effectiveness.

@@

The surveillance dataset is structured to conform with state and federal reporting requirements. As a result, the data format must be transformed to make it ready for analysis. Once the data structure is corrected, it can be joined to other relevant datasets using spatial or temporal attributes.

@@

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold or degree-days. Degree-days is a measure of the amount of
heat accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006]. To simplify this
variable, the number of hours the temperature was ≥50°F and ≤86°F. These
thresholds were chosen based on the known temperature range for mosquito
development and virus transmission.

@@

In response to the observed non-linear relationships in the data, a
two-stage Generalized Additive Model (GAM) analysis was used. GAMs
provide a versatile framework for capturing complex, non-linear
relationships between response variables and multiple predictors, often encountered in ecological data sets.
